---
title: "Descrip_lutuk"
author: "Alfredo Acosta"
date: "2024-05-26"
output:
  html_document: 
    toc: yes
    toc_float: true
    theme: cosmo
    fig_caption: yes
    number_sections: yes
    code_folding: hide
  word_document: default
  html_notebook: 
    toc: yes
---
# Survey Mapping ASF outbreak Lutuk March 2024
Instructions: 
Go to all households in Lutuk that have pigs or that had pigs at the start of this outbreak
Ask if there is anyone at home who is involved with the pigs. If not, go back later. 
# Organizing dataset
## Loading libraries 
```{r }
# Library ----
library(readxl)
library(dplyr)
library(lubridate)
library(leaflet)
library(sf)
library(DT)
library(psych)
library(sjmisc)
library(ggplot2)
```
## Working directory
```{r}
setwd("C:/Users/alfredo.acosta/OneDrive - SVA/ASF_Uganda/R/")
```

## Reading excel file
```{r}
l <- read_xlsx("Kopia av Lutuk_data collection on ASF outbreak 2024 MIRED May 6.xlsx")
colnames(l)
```

## Colnames preparation

```{r}
colnames(l) <- c("1digited_by",
"l2_Way_point",                                                       
"l3_NGwenAna_col",                                                            
"l4_Sub_village",                                                                             "l5_Date_interview",                                                                        
"l6_Name_producer",
"l7_Gender",
"l8_Age_years",
"l9_Norting",
"l10_Easting",
"l11_Interviewed_by_G&Ana",                                                                  
"l12_first_detection_ill_pig",                                    
"l13_first_pig_die",                                                    
"l14_N_init",                               
"l15_N_affected",              
"l16_N_deaths",                                                                              
"l17_N_interview",                                                                   
"l18_pnd_slaughtered",#Pigsnot dead (pnd)                                   
"l19_pnd_sold_alive",                                   
"l20_pnd_kept",                                 
"l21_pnd_other",                                        
"l22_pnd_other(what)",                                       
"l23_pd_Ate",
"l24_pd_Sold",
"l25_pd_buried",
"l26_pd_burned",
"l27_pd_latrine",
"l28_pd_bush_swamp_river",
"l29_pd_other",
"l30_pd_other_what",
"l31_Why_sick_Weather",
"l32_Why_Witchcraft",
"l33_Why_Poison",
"l34_Why_Jealousy",
"l35_Why_Someone_killed_it",
"l36_Why_pig_ate_something",
"l37_Why_infected_by_sick_pig", 
"l38_Why_got_infected_by_a_person",
"l39_Why_Other",
"l40_Why_(reason_other)",
"l41_If_your_pig_was_infected_by_another_pig_or_person,how_this happened?",
"l42_What_do_Slaughtered",              
"l43_What_do_Sold_alive", 
"l44_What_do_Slaughtered_healthy_pig(s)",           
"l45_What_do_Sold_healthy_pig",        
"l46_What_Prayed_",
"l47_What_Separated_ill_from",     
"l48_What_do_tied_my_pigs_fenced",        
"l49_What_do_Gave_medication",
"l50_What_do_(what)",
"l51_What_do_asked_for_help",
"l52_What_do_asked_for_help(whom)",
"l53_What_do_Other_(yes/no)",
"l54_What_do_Other(what)",
"l55_How_this_affects_your_life?",
"l56_How_do_you_feel",
"l57_Do_you_think_invest_again",               
"l58_Why_or_why_not?",                                                                       "l59_Is there anything you like to add?",                                                    "l60_Kommentarer från den som fyller i Excel")
```

```{r}
colnames(l)
```
# When did you first detect and ill pig in this outbreak
## Date checking
```{r}
l$Date_first_detection_ill_pig <- dmy(l$l12_first_detection_ill_pig)
l$Date_first_pig_die <- dmy(l$l13_first_pig_die)
l$Date_interview <- dmy(l$l5_Date_interview)

l %>% 
  select(l6_Name_producer, l12_first_detection_ill_pig,
         Date_first_detection_ill_pig, l16_N_deaths) %>% 
  filter(is.na(Date_first_detection_ill_pig) & l16_N_deaths > 0)
```
## Checking temporal evolution
## Two surveys withouth date of first detection even though there is dead pigs
## ? (please check two surveys Anywar Cody, Aryema Stephen)
```{r}
l %>% 
group_by(l6_Name_producer) %>% 
  filter(is.na(l12_first_detection_ill_pig) & l16_N_deaths > 0)
```

## ? Please check this surveys where there is not date of first pig deaths ()
```{r}
l %>% 
select(l5_Date_interview,
       l6_Name_producer,
       Date_first_pig_die,
       l16_N_deaths) %>% 
  group_by(l6_Name_producer) %>% 
  filter(is.na(Date_first_pig_die) & l16_N_deaths) %>% 
  filter(is.na(Date_first_pig_die))
```

## 6 cases in 2023 and 46 in 2024 NA 47
```{r}
l %>% 
  group_by(year(l$Date_first_detection_ill_pig)) %>% 
  summarise(n())

```

```{r}
l %>% 
  group_by(year(l$Date_first_detection_ill_pig),
           month(l$Date_first_detection_ill_pig, label=TRUE)) %>% 
  summarise(n())


```
```{r}
l %>% 
  group_by(Month=month(Date_first_detection_ill_pig)) %>% 
  summarise(outbreaks=n()) %>% 
  ggplot(aes(Month, outbreaks))+
  geom_col()

```
```{r}
l %>% 
  group_by(Month=floor_date(Date_first_detection_ill_pig, unit="month")) %>% 
  summarise(outbreaks=n()) %>% 
  ggplot(aes(Month, outbreaks))+
  geom_col()

```
## time differences
## ? Olinga David Paska, the date of interview is precedent from the first detection of ill pigs, check the dates.
## ? Agen, Oling, Opiyo, there is not death pigs but still the interview was made in the past 7, 278 and 1 days before the first detection.
```{r}

l$dif_first_detection_to_interview <- l$Date_interview - l$Date_first_detection_ill_pig

summary(as.numeric(l$Date_interview - l$Date_first_detection_ill_pig))

l %>% 
  group_by(l6_Name_producer, 
           l5_Date_interview, 
           l12_first_detection_ill_pig, 
           l13_first_pig_die) %>% 
  filter(dif_first_detection_to_interview <0) %>%
  summarise(n(), sum(dif_first_detection_to_interview))
```

```{r}
l %>% 
  group_by(l6_Name_producer, 
           l5_Date_interview, 
           l12_first_detection_ill_pig, 
           l13_first_pig_die) %>% 
  filter(dif_first_detection_to_interview >0) %>%
  summarise(n(), sum(dif_first_detection_to_interview))

l %>% 
  filter(dif_first_detection_to_interview >0) %>%
  summarise(n(), mean(dif_first_detection_to_interview))

l %>% 
  group_by(l4_Sub_village) %>% 
  filter(dif_first_detection_to_interview >0) %>%
  summarise(n(), interval=mean(dif_first_detection_to_interview)) %>% 
  arrange(desc(interval))


```


# Fixing grouped variables
```{r}
l$l14_N_init <- as.numeric(l$l14_N_init)
l$l15_N_affected <- as.numeric(l$l15_N_affected)
l$l16_N_deaths <- as.numeric(l$l16_N_deaths)
l$l17_N_interview <- as.numeric(l$l17_N_interview)

#pigs not death
l$l18_pnd_slaughtered <- gsub("No", "0", l$l18_pnd_slaughtered)
l$l18_pnd_slaughtered <- as.numeric(l$l18_pnd_slaughtered)

l$l19_pnd_sold_alive <- gsub("No", "0", l$l19_pnd_sold_alive)
l$l19_pnd_sold_alive <- as.numeric(l$l19_pnd_sold_alive)

l$l20_pnd_kept <- as.numeric(l$l20_pnd_kept)

l$l19_pnd_sold_alive <- as.numeric(l$l21_pnd_other)
```
# Population check
## Individual farms were balance is not 0 (death+non.death(slaughter+sold+kept)). Please check this 16 surveys:
i.e. Owor Gedge at the start of the outbreak had 20 pigs, no deaths but kept 2, maybe he kept 20 instead of 2. The balance is 18. 20-2=18
Apiyo Grace at the start of the outbreak had 1 pig, no deaths, he kept 10, maybe he kept 1 instead of 10. Balance is -9. 1-10=-9.
```{r}
datatable(l %>% 
  # group_by(l4_Sub_village) %>% 
  group_by(l6_Name_producer) %>% 
  summarise(n(),Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            # Affected=sum(l15_N_affected, na.rm = TRUE),
            Not_dead=sum(l18_pnd_slaughtered, 
                         l19_pnd_sold_alive, 
                         l20_pnd_kept, na.rm = TRUE),
            Balance=Initial_pop-Deaths-Not_dead,
            # start=Initial_pop - Now,
    slaughther=sum(l18_pnd_slaughtered, na.rm = TRUE),
            Sold=sum(l19_pnd_sold_alive, na.rm = TRUE),
            Kept=sum(l20_pnd_kept, na.rm = TRUE),
            Interview=sum(l17_N_interview, na.rm = TRUE),
) %>% 
  arrange(desc(Balance)) %>% 
  filter(Balance != 0) %>% 
  print(n=100))


```

## Individual farms were balance is 0 (death+non.death(slaughter+sold+kept)). 83 entries are ok. Balance is checked considering the number of initial pigs at the beggining of the outbreak minus the slaughtered, sold or kept.
```{r}
datatable(l %>% 
  # group_by(l4_Sub_village) %>% 
  group_by(l6_Name_producer) %>% 
  summarise(n(),Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            # Affected=sum(l15_N_affected, na.rm = TRUE),
            Not_dead=sum(l18_pnd_slaughtered, 
                         l19_pnd_sold_alive, 
                         l20_pnd_kept, na.rm = TRUE),
            Balance=Initial_pop-Deaths-Not_dead,
            # start=Initial_pop - Now,
    slaughther=sum(l18_pnd_slaughtered, na.rm = TRUE),
            Sold=sum(l19_pnd_sold_alive, na.rm = TRUE),
            Kept=sum(l20_pnd_kept, na.rm = TRUE),
            # Interview=sum(l17_N_interview, na.rm = TRUE),
) %>% 
  arrange(desc(Balance)) %>% 
  filter(Balance == 0) %>% 
  print(n=100))
```
## Balance check grouped by village.
```{r}
datatable(l %>% 
  group_by(l4_Sub_village) %>%
  # group_by(l6_Name_producer) %>% 
  summarise(n(),Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            # Affected=sum(l15_N_affected, na.rm = TRUE),
            Not_dead=sum(l18_pnd_slaughtered, 
                         l19_pnd_sold_alive, 
                         l20_pnd_kept, na.rm = TRUE),
            Balance=Initial_pop-Deaths-Not_dead,
            # start=Initial_pop - Now,
    slaughther=sum(l18_pnd_slaughtered, na.rm = TRUE),
            Sold=sum(l19_pnd_sold_alive, na.rm = TRUE),
            Kept=sum(l20_pnd_kept, na.rm = TRUE),
            # Interview=sum(l17_N_interview, na.rm = TRUE),
) %>% 
  arrange(desc(Balance)) %>% 
  # filter(Balance == 0) %>% 
  print(n=100))
```

## Calculating aparent prevalence 
```{r}
l$l61_apparent_prevalence <- l$l16_N_deaths/l$l14_N_init
l$l62attack_rate <- l$l16_N_deaths/l$l14_N_init

l %>% 
  select(l14_N_init, 
         l15_N_affected, 
         l16_N_deaths, 
         l18_pnd_slaughtered, 
         l19_pnd_sold_alive, 
         l20_pnd_kept,
         l17_N_interview,
         l61_apparent_prevalence) %>% 
  descr()

summary(l$l61_apparent_prevalence)
```
```{r}
l %>% 
  group_by(l4_Sub_village) %>% 
  select(l14_N_init, 
         l15_N_affected, 
         l16_N_deaths, 
         l18_pnd_slaughtered, 
         l19_pnd_sold_alive, 
         l20_pnd_kept,
         l17_N_interview,
         l61_apparent_prevalence) %>% 
  descr()

```

# Apparent prevalence by farm
```{r}
datatable(l %>% 
  # group_by(l4_Sub_village) %>%
  group_by(l6_Name_producer) %>%
  summarise(n(), Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            # Affected=sum(l15_N_affected, na.rm = TRUE),
            Apparent_prevalence=round((Deaths/Initial_pop),4)*100) %>% 
  arrange(desc(Apparent_prevalence)) %>% 
  # filter(Balance == 0) %>% 
  print(n=100))

```

# Total Apparent prevalence by village
```{r}
datatable(l %>% 
  group_by(l4_Sub_village) %>%
  summarise(n(), Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            # Affected=sum(l15_N_affected, na.rm = TRUE),
            Apparent_prevalence=round((Deaths/Initial_pop),4)*100) %>% 
  arrange(desc(Apparent_prevalence)) %>% 
  # filter(Balance == 0) %>% 
  print(n=100))

```


## Distribution of initial and final population over villages
```{r}
ggplot(l, aes(l14_N_init, l4_Sub_village, fill=l4_Sub_village))+
  geom_boxplot()+
  labs(
    y=NULL,
    x="Number of pigs at the begining of the outbreak"
  )

ggplot(l, aes(l17_N_interview, l4_Sub_village, fill=l4_Sub_village))+
  geom_boxplot()+
  labs(
    y=NULL,
    x="Number of pigs at the interview"
  )

```
# How many pigs did the household have at the start of the outbreak


# What did you do with the pigs that died?
## Check errors of problems (nothing wrong found)
```{r}
l$l23_pd_Ate[l$l23_pd_Ate == "NA"] <- NA
l$l24_pd_Sold[l$l24_pd_Sold == "NA"] <- NA
l$l25_pd_buried[l$l25_pd_buried == "NA"] <- NA
l$l26_pd_burned[l$l26_pd_burned == "NA"] <- NA
l$l27_pd_latrine[l$l27_pd_latrine == "NA"] <- NA
l$l28_pd_bush_swamp_river[l$l28 == "NA"] <- NA

l %>% 
  filter(is.na(l23_pd_Ate) & l16_N_deaths != 0)

l %>% 
  filter(is.na(l24_pd_Sold) & l16_N_deaths != 0)

l %>% 
  filter(is.na(l25_pd_buried) & l16_N_deaths != 0)

l %>% 
  filter(is.na(l26_pd_burned) & l16_N_deaths != 0)

l %>% 
  filter(is.na(l27_pd_latrine) & l16_N_deaths != 0)

```
## Changing to numerical to quantify
```{r}
l$pd_ate <- l$l23_pd_Ate
l$pd_ate <- gsub("Yes", "1", l$pd_ate)
l$pd_ate <- gsub("No", "0", l$pd_ate)

l$pd_sold <- l$l24_pd_Sold
l$pd_sold <- gsub("Yes", "1", l$pd_sold)
l$pd_sold <- gsub("No", "0", l$pd_sold)

l$pd_buried <- l$l25_pd_buried
l$pd_buried <- gsub("Yes", "1", l$pd_buried)
l$pd_buried <- gsub("No", "0", l$pd_buried)

l$pd_burned <- l$l26_pd_burned
l$pd_burned <- gsub("Yes", "1", l$pd_burned)
l$pd_burned <- gsub("No", "0", l$pd_burned)

l$pd_latrine <- l$l27_pd_latrine
l$pd_latrine <- gsub("Yes", "1", l$pd_latrine)
l$pd_latrine <- gsub("No", "0", l$pd_latrine)

l$pd_bush_river <- l$l28_pd_bush_swamp_river
l$pd_bush_river <- gsub("Yes", "1", l$pd_bush_river)
l$pd_bush_river <- gsub("No", "0", l$pd_bush_river)

l$pd_other <- l$l29_pd_other
l$pd_other <- gsub("Yes", "1", l$pd_other)
l$pd_other <- gsub("No", "0", l$pd_other)

l$pd_ate <- as.numeric(l$pd_ate)
l$pd_sold <- as.numeric(l$pd_sold)
l$pd_buried <- as.numeric(l$pd_buried)
l$pd_burned <- as.numeric(l$pd_burned)
l$pd_latrine <- as.numeric(l$pd_latrine)
l$pd_bush_river <- as.numeric(l$pd_bush_river)
l$pd_other <- as.numeric(l$pd_ate)

```
```{r}

l %>% 
  # group_by(l$l4_Sub_village) %>%
  summarise(Ate=sum(pd_ate, na.rm =TRUE),
            Sold=sum(pd_sold, na.rm =TRUE),
            Burried=sum(pd_buried, na.rm =TRUE),
            Burned=sum(pd_burned, na.rm =TRUE),
            Latrine=sum(pd_latrine, na.rm =TRUE),
            Bush=sum(pd_bush_river, na.rm =TRUE),
            Other=sum(pd_other, na.rm =TRUE),)

l %>% 
  group_by(l$l4_Sub_village) %>%
  summarise(Ate=sum(pd_ate, na.rm =TRUE),
            Sold=sum(pd_sold, na.rm =TRUE),
            Burried=sum(pd_buried, na.rm =TRUE),
            Burned=sum(pd_burned, na.rm =TRUE),
            Latrine=sum(pd_latrine, na.rm =TRUE),
            Bush=sum(pd_bush_river, na.rm =TRUE),
            Other=sum(pd_other, na.rm =TRUE),)

l$l30_pd_other_what

```

<!-- # # What do you think caused the pig to get sick? (you can pick several)  -->
<!-- # ```{r} -->
<!-- # ## Changing to numerical to quantify -->
<!-- # ```{r} -->
<!-- #  -->
<!-- # l$l31_Why_sick_Weather[l$l31_Why_sick_Weather == "NA"] <- NA -->
<!-- # l$l32_Why_Witchcraft[l$l32_Why_Witchcraft == "NA"] <- NA -->
<!-- # l$l33_Why_Poison[l$l33_Why_Poison == "NA"] <- NA -->
<!-- # l$l34_Why_Jealousy[l$l34_Why_Jealousy == "NA"] <- NA -->
<!-- # l$l34_Why_Jealousy[l$l35_Why_Someone_killed_it == "NA"] <- NA -->
<!-- # l$l35_Why_Someone_killed_it[l$l35_Why_Someone_killed_it == "NA"] <- NA -->
<!-- # l$l36_Why_pig_ate_something[l$l36_Why_pig_ate_something == "NA"] <- NA -->
<!-- # l$l37_Why_infected_by_sick_pig[l$l37_Why_infected_by_sick_pig== "NA"] <- NA -->
<!-- # l$l38_Why_got_infected_by_a_person[l$l38_Why_got_infected_by_a_person == "NA"] <-  -->
<!-- #  -->
<!-- #  -->
<!-- # l$why_sick_weather <- l[,2] -->
<!-- # l$why_sick_weater <- gsub("Yes", "1", l$why_sick_weater) -->
<!-- # l$pd_weather <- gsub("No", "0", l$pd_weather) -->
<!-- #  -->
<!-- # l$pd_sold <- l$l24_pd_Sold -->
<!-- # l$pd_sold <- gsub("Yes", "1", l$pd_sold) -->
<!-- # l$pd_sold <- gsub("No", "0", l$pd_sold) -->
<!-- #  -->
<!-- # l$pd_buried <- l$l25_pd_buried -->
<!-- # l$pd_buried <- gsub("Yes", "1", l$pd_buried) -->
<!-- # l$pd_buried <- gsub("No", "0", l$pd_buried) -->
<!-- #  -->
<!-- # l$pd_burned <- l$l26_pd_burned -->
<!-- # l$pd_burned <- gsub("Yes", "1", l$pd_burned) -->
<!-- # l$pd_burned <- gsub("No", "0", l$pd_burned) -->
<!-- #  -->
<!-- # l$pd_latrine <- l$l27_pd_latrine -->
<!-- # l$pd_latrine <- gsub("Yes", "1", l$pd_latrine) -->
<!-- # l$pd_latrine <- gsub("No", "0", l$pd_latrine) -->
<!-- #  -->
<!-- # l$pd_bush_river <- l$l28_pd_bush_swamp_river -->
<!-- # l$pd_bush_river <- gsub("Yes", "1", l$pd_bush_river) -->
<!-- # l$pd_bush_river <- gsub("No", "0", l$pd_bush_river) -->
<!-- #  -->
<!-- # l$pd_other <- l$l29_pd_other -->
<!-- # l$pd_other <- gsub("Yes", "1", l$pd_other) -->
<!-- # l$pd_other <- gsub("No", "0", l$pd_other) -->
<!-- #  -->
<!-- # l$pd_ate <- as.numeric(l$pd_ate) -->
<!-- # l$pd_sold <- as.numeric(l$pd_sold) -->
<!-- # l$pd_buried <- as.numeric(l$pd_buried) -->
<!-- # l$pd_burned <- as.numeric(l$pd_burned) -->
<!-- # l$pd_latrine <- as.numeric(l$pd_latrine) -->
<!-- # l$pd_bush_river <- as.numeric(l$pd_bush_river) -->
<!-- # l$pd_other <- as.numeric(l$pd_ate) -->

# Map
## Preparing coordinates

## ? 	Check Acayo Nighthy Santo nort coordinate (y), it is the same as x. Correct.
## ? Check Christine Lukioiya / Innocen atimango easting (x) coordinate

```{r}
l$y <- l$l9_Norting
l$y <- gsub("-", ".", l$y)
l$y <- as.numeric(l$y)

l$x <- l$l10_Easting
l$x <- gsub("-", ".", l$x)
l$x <- as.numeric(l$x)

plot(l$x, l$y)

ggplot

```

## Manipulable map of the complete dataset
```{r}
library(leaflet)
l$owner <- l$l6_Name_producer

map <- leaflet(l) %>%
  addProviderTiles(providers$Esri)

mc <- map %>%
  setView(lng = 32, lat = 2, zoom = 7) %>%
  addMarkers(~x, ~y,
              popup = ~paste("Owner:", owner,
                             "<br>", "Prevalence", l61_apparent_prevalence,
                             "<br>", "Deaths", l16_N_deaths))

mc

```

## Manipulable map of the cases (farms with reported deaths)
```{r}

l2 <- l %>% 
  filter(l16_N_deaths > 0)


map2 <- leaflet(l2) %>%
  # addProviderTiles(providers$OpenStreetMap)
  addProviderTiles(providers$Esri.WorldImagery)

mp <- map2 %>%
  setView(lng = 32, lat = 2, zoom = 7) %>%
  addMarkers(~x, ~y,
              popup = ~paste("Owner:", owner,
                             "<br>", "Aparent Prev", l61_apparent_prevalence,
                             "<br>", "Deaths", l16_N_deaths))

mp

```

# Credits 
Acosta, Alfredo PhD<sup>1</sup>. 
**SVA<sup>1</sup>**: SVA <http://www.sva.se/>. 