---
title: "Descrip_lutuk"
author: "Alfredo Acosta"
date: "2024-05-26"
output:
  html_document: 
    toc: yes
    toc_float: true
    theme: cosmo
    fig_caption: yes
    number_sections: yes
    code_folding: hide
  word_document: default
  html_notebook: 
    toc: yes
---
# Survey Mapping ASF outbreak Lutuk March 2024
Instructions: 
Go to all households in Lutuk that have pigs or that had pigs at the start of this outbreak
Ask if there is anyone at home who is involved with the pigs. If not, go back later. 
# Organizing dataset
## Loading libraries 
```{r }
# Library ----
library(readxl)
library(arsenal)
library(dplyr)
library(lubridate)
library(leaflet)
library(sf)
library(DT)
library(psych)
library(sjmisc)
library(ggplot2)
```
## Working directory
```{r}
setwd("C:/Users/alfredo.acosta/OneDrive - SVA/ASF_Uganda/R/")
```

## Reading excel file
```{r}
l <- read_xlsx("Kopia av Lutuk_data collection on ASF outbreak 2024 MIRED May 6.xlsx", col_types = "text")
m <- read_xlsx("kopia av Lutuk_data collection on ASF outbreak 2024 MIRED May 6 GV.xlsx", col_types = "text")
```
# Comparison between versions 
```{r}
comparedf(l, m)
summary(comparedf(l, m))

# Differences detected in administration variables. deleting m
rm(m)
```
 
## Colnames preparation
```{r}
colnames(l) <- c("1digited_by",
"l2_Way_point",                                                       
"l3_NGwenAna_col",                                                            
"l4_Sub_village",
"l5_Date_interview",                                                                        
"l6_Name_producer",
"l7_Gender",
"l8_Age_years",
"l9_Norting",
"l10_Easting",
"l11_Interviewed_by_G&Ana",                                                                  
"l12_first_detection_ill_pig",                                    
"l13_first_pig_die",                                                    
"l14_N_init",                               
"l15_N_affected",              
"l16_N_deaths",                                                                              
"l17_N_interview",                                                                   
"l18_pnd_slaughtered",#Pigsnot dead (pnd)                                   
"l19_pnd_sold_alive",                                   
"l20_pnd_kept",                                 
"l21_pnd_other",                                        
"l22_pnd_other(what)",                                       
"l23_pd_Ate",
"l24_pd_Sold",
"l25_pd_buried",
"l26_pd_burned",
"l27_pd_latrine",
"l28_pd_bush_swamp_river",
"l29_pd_other",
"l30_pd_other_what",
"l31_Why_sick_Weather",
"l32_Why_Witchcraft",
"l33_Why_Poison",
"l34_Why_Jealousy",
"l35_Why_Someone_killed_it",
"l36_Why_pig_ate_something",
"l37_Why_infected_by_sick_pig", 
"l38_Why_got_infected_by_a_person",
"l39_Why_Other",
"l40_Why_(reason_other)",
"l41_If_your_pig_was_infected_by_another_pig_or_person,how_this happened?",
"l42_What_do_Slaughtered",              
"l43_What_do_Sold_alive", 
"l44_What_do_Slaughtered_healthy_pig(s)",           
"l45_What_do_Sold_healthy_pig",        
"l46_What_Prayed_",
"l47_What_Separated_ill_from",     
"l48_What_do_tied_my_pigs_fenced",        
"l49_What_do_Gave_medication",
"l50_What_do_(what)",
"l51_What_do_asked_for_help",
"l52_What_do_asked_for_help(whom)",
"l53_What_do_Other_(yes/no)",
"l54_What_do_Other(what)",
"l55_How_this_affects_your_life?",
"l56_How_do_you_feel",
"l57_Do_you_think_invest_again",               
"l58_Why_or_why_not?",
"l59_Is there anything you like to add?",
"l60_Kommentarer från den som fyller i Excel")
```

```{r}
colnames(l)
```
# Date checking
## Premises with death pigs but without date of first detection of ill pig
## ? 66 and 116 excluded
```{r}
l$Date_first_detection_ill_pig <- dmy(l$l12_first_detection_ill_pig)
l$Date_first_pig_die <- dmy(l$l13_first_pig_die)
l$Date_interview <- dmy(l$l5_Date_interview)

# l %>% 
#   select(l2_Way_point, l6_Name_producer, 
#          Date_interview,
#          l12_first_detection_ill_pig,
#          Date_first_detection_ill_pig, l16_N_deaths) %>% 
#   filter(is.na(Date_first_detection_ill_pig) & l16_N_deaths > 0)

#Correcting 66 from the form 
l$Date_first_detection_ill_pig[l$l2_Way_point == 66] <- l$Date_interview[l$l2_Way_point == 66] - days(53)
l$Date_first_detection_ill_pig[l$l2_Way_point == 116] <- l$Date_interview[l$l2_Way_point == 116] - days(53)
```
## Premises with death pigs but without date of first death of pig
## ?? 37,38,40,42,43 I assume that date of first pig die will be the mean days between detection and die 
```{r eval=FALSE, include=FALSE}
l %>% 
select(l2_Way_point,
       l5_Date_interview,
       l6_Name_producer,
       Date_first_detection_ill_pig,
       Date_first_pig_die,
       l16_N_deaths) %>% 
  filter(is.na(Date_first_pig_die) & as.numeric(l16_N_deaths) >0)
```

## 6 outbreaks in 2023 and 46 in 2024 NA 47
```{r eval=FALSE, include=FALSE}
l %>% 
  group_by(year(Date_first_detection_ill_pig)) %>% 
  summarise(n())

l %>% 
  group_by(year(l$Date_first_pig_die)) %>% 
  summarise(n())

```

## Replacing 2023 to 2024 in first detection and death
## 31,37,46,49,72,79 replaced date of first detection in 2023 to 2024 interview date and death date matches
## ?31 replaced date of death 2024 to 2023 
```{r}
# l %>% 
#   filter(year(Date_first_detection_ill_pig) == 2023) %>% 
#   select(l2_Way_point,
#        l6_Name_producer,
#        Date_first_detection_ill_pig,
#        Date_interview,
#        Date_first_pig_die,
#        l16_N_deaths)

l <- l %>% 
  mutate(Date_first_detection_ill_pig = as.Date(ifelse(lubridate::year(Date_first_detection_ill_pig) == 2023, 
         Date_first_detection_ill_pig + lubridate::years(1),
         Date_first_detection_ill_pig)))

# l %>% 
#   filter(year(Date_first_pig_die) == 2023) %>% 
#   select(l2_Way_point,
#        l6_Name_producer,
#        Date_first_detection_ill_pig,
#        Date_interview,
#        Date_first_pig_die,
#        l16_N_deaths)

l <- l %>% 
  mutate(Date_first_pig_die = as.Date(ifelse(lubridate::year(Date_first_pig_die) == 2023, 
         Date_first_pig_die + lubridate::years(1),
         Date_first_pig_die)))

l <- l %>% 
  mutate(Date_first_detection_ill_pig = as.Date(ifelse(lubridate::year(Date_first_detection_ill_pig) != 2024, 
         Date_first_detection_ill_pig + lubridate::years(-1),
         Date_first_detection_ill_pig)))

```

## Outbreaks with date of first_detection inoctober and december
```{r}
l %>% 
  group_by(year(l$Date_first_detection_ill_pig),
           month(l$Date_first_detection_ill_pig, label=TRUE)) %>% 
  summarise(n())

l %>% 
  group_by(year(l$Date_first_pig_die),
           month(l$Date_first_pig_die, label=TRUE)) %>% 
  summarise(n())

```
```{r}
l %>% 
  group_by(Month=month(Date_first_detection_ill_pig)) %>% 
  summarise(outbreaks=n()) %>% 
  ggplot(aes(Month, outbreaks))+
  geom_col()+
  labs(x="Date of first detection ill pigs")

l %>% 
  group_by(Month=month(Date_first_pig_die)) %>% 
  summarise(outbreaks=n()) %>% 
  ggplot(aes(Month, outbreaks))+
  geom_col()+
  labs(x="Date of first pig die")

```
## Events in October and december, make sense??
## form 98 corrected Date of interview from image
```{r}
# l %>% 
# select(l2_Way_point,
#        l5_Date_interview,
#        Date_interview,
#        l6_Name_producer,
#        Date_first_detection_ill_pig,
#        Date_first_pig_die,
#        l16_N_deaths) %>% 
#   filter(month(Date_first_detection_ill_pig) >5) 

#Correcting 98 from the form 
l$Date_interview[l$l2_Way_point == 98] <- as.Date("2024-03-27")

#Correcting 67 and 98 from 2024 to 2023
l <- l %>% 
  mutate(Date_first_detection_ill_pig = as.Date(ifelse(lubridate::month(Date_first_detection_ill_pig) > 5, 
         Date_first_detection_ill_pig + lubridate::years(-1),
         Date_first_detection_ill_pig)))


# l %>% 
# select(l2_Way_point,
#        l5_Date_interview,
#        Date_interview,
#        l6_Name_producer,
#        Date_first_detection_ill_pig,
#        Date_first_pig_die,
#        l16_N_deaths) %>% 
#   filter(month(Date_first_pig_die) >5)


l <- l %>% 
  mutate(Date_first_pig_die = as.Date(ifelse(lubridate::month(Date_first_pig_die) > 5, 
         Date_first_pig_die + lubridate::years(-1),
         Date_first_pig_die)))


```
# Visualization of months
```{r eval=FALSE, include=FALSE}
l %>% 
  group_by(month=floor_date(Date_first_detection_ill_pig, unit= "month")) %>% 
  summarise(outbreaks=n()) %>% 
  ggplot(aes(month, outbreaks))+
  geom_col()+
  geom_text(aes(label=outbreaks), vjust= -0.5)+
  ylim(0,35)+
  labs(x="Date of first detection ill pigs")

l %>% 
  group_by(month=floor_date(Date_first_pig_die, unit= "month")) %>% 
  summarise(outbreaks=n()) %>% 
  ggplot(aes(month, outbreaks))+
  geom_col()+
  geom_text(aes(label=outbreaks), vjust= -0.5)+
  ylim(0,30)+
  labs(x="Date of first pig is dead")

```

## Time differences
## 101, 40 thefirst detection of ill pigs happens after the interview (7 and 1 days), check the dates.
## 35 first pig die before detecting the disease. The problem is incorrect transcript first detection 3 or 5 march, first pig die on 7 of march and the interview was on 15 of march. Corrected.
```{r}
l$dif_first_detection_to_interview <- l$Date_interview - l$Date_first_detection_ill_pig

l$dif_first_pig_die_to_interview <- l$Date_interview - l$Date_first_pig_die

summary(as.numeric(l$dif_first_detection_to_interview))

summary(as.numeric(l$dif_first_detection_to_interview))
summary(as.numeric(l$dif_first_pig_die_to_interview))

boxplot(as.numeric(l$dif_first_detection_to_interview))
boxplot(as.numeric(l$dif_first_pig_die_to_interview))

l$l16_N_deaths <- as.numeric(l$l16_N_deaths)
summary(as.numeric(l$Date_interview - l$Date_first_detection_ill_pig))

l %>% 
  group_by(l4_Sub_village) %>% 
  summarise(n(), mean(dif_first_detection_to_interview, na.rm = TRUE))



# l %>% 
#   group_by(l2_Way_point,
#           l6_Name_producer, 
#            Date_first_detection_ill_pig, 
#            Date_first_pig_die,
#            Date_interview, 
#            dif_first_detection_to_interview) %>% 
#   filter(dif_first_detection_to_interview <0) %>%
#   summarise(n())

#Correcting 35 from the form 
l$Date_first_detection_ill_pig[l$l2_Way_point == 35] <- as.Date("2024-03-05")

```

## Time difference between detection and death
## 116 
## 80 The first pig died 19 days after detecting the first pig ill. It makes sense if the 9/02/2024 is 29/02/2024 because the first detected ill was on 28/02/2024 (corrected) 
```{r}
l$dif_detection_death <- l$Date_first_pig_die - l$Date_first_detection_ill_pig

summary(as.numeric(l$dif_detection_death))

boxplot(as.numeric(l$dif_detection_death))


# l %>% 
#   group_by(l2_Way_point,
#           l6_Name_producer, 
#            Date_first_detection_ill_pig, 
#            Date_first_pig_die,
#            Date_interview, 
#            dif_first_detection_to_interview,
#             dif_detection_death) %>% 
#   filter(l16_N_deaths >0) %>%
#   summarise(n()) %>% 
#   arrange(desc(dif_detection_death))

# Corrected 80
l$Date_first_pig_die[l$l2_Way_point == 80] <- as.Date("2024-02-29")


```

## Selection of cases=1, controls=0
```{r}
l %>% 
  group_by(l2_Way_point,
           Date_first_detection_ill_pig, 
           Date_first_pig_die,
           Date_interview, 
           dif_first_detection_to_interview,
          dif_detection_death) %>% 
  filter(l16_N_deaths >0) %>%
  filter(!is.na(Date_first_detection_ill_pig)) %>% 
  summarise(n()) %>% 
  arrange(desc(dif_detection_death))

l <- l %>%
  mutate(cases = ifelse(!is.na(Date_first_detection_ill_pig) & l16_N_deaths > 0, "Case", "Control"))

table(l$cases)


# l %>% 
#   select(l2_Way_point, cases, l16_N_deaths, l14_N_init, l15_N_affected) %>% 
#   filter(cases == "Control") %>% 
#   filter(l16_N_deaths >0)
# 
# l %>% 
#   select(l2_Way_point, cases, l16_N_deaths, l14_N_init, l15_N_affected, Date_first_detection_ill_pig, Date_first_pig_die) %>% 
#   filter(cases == "Control") %>% 
#   filter(year(Date_first_detection_ill_pig) ==  "2023")


```
# Graph epidemiologcal presentation

```{r}

datatable(l %>%
  group_by(cases, week=floor_date(Date_first_detection_ill_pig, unit= 'week'))%>%
  filter(cases == "Case") %>% 
  # filter(week > "2024-01-01") %>% 
  summarise(Cases=sum(l16_N_deaths),
            Outbreaks=sum(cases == 'Case')) %>%
  arrange(week))

l %>%
  group_by(cases, week=floor_date(Date_first_detection_ill_pig, unit= 'week'))%>%
  filter(cases == "Case") %>% 
  # filter(week > "2024-01-01") %>% 
  summarise(Cases=sum(l16_N_deaths),
            Outbreaks=sum(cases == 'Case'),
            Observations=n()) %>%  
  ggplot()+
  geom_col(aes(week, Cases), fill="#FC8D62")+
  # geom_col(aes(week, Observations), fill="#21918c") +
  geom_point(aes(week, Outbreaks*2), size=3, color="#A6D854")+
  geom_line(aes(week, Outbreaks*2), size=0.1, color="#A6D854")+
  scale_y_continuous(
  sec.axis = sec_axis(trans = ~. /2, name="Number of outbreaks")) +
  labs(fill="",
       x="Epidemiological week",
       y="Number of death pigs")+
  theme_minimal()+
  theme(text = element_text(size = 14)) +
  scale_color_brewer(palette = "Set2")

# brewer.pal(n = 4, name = "Set1")

```
## Deleting cases
```{r}
l <- l[l$l2_Way_point != "66",]
l <- l[l$l2_Way_point != "67",]
l <- l[l$l2_Way_point != "116",]
```



```{r}
l %>% 
  filter(dif_first_detection_to_interview >0) %>%
  summarise(n(), mean(dif_first_detection_to_interview))

l %>% 
  group_by(l4_Sub_village) %>% 
  filter(dif_first_detection_to_interview >0) %>%
  summarise(n(), interval=mean(dif_first_detection_to_interview)) %>% 
  arrange(desc(interval))
```

# Fixing grouped variables
```{r}
l$l14_N_init <- as.numeric(l$l14_N_init)
l$l15_N_affected <- as.numeric(l$l15_N_affected)
l$l16_N_deaths <- as.numeric(l$l16_N_deaths)
l$l17_N_interview <- as.numeric(l$l17_N_interview)

#pigs not death
l$l18_pnd_slaughtered <- gsub("No", "0", l$l18_pnd_slaughtered)
l$l18_pnd_slaughtered <- as.numeric(l$l18_pnd_slaughtered)

l$l19_pnd_sold_alive <- gsub("No", "0", l$l19_pnd_sold_alive)
l$l19_pnd_sold_alive <- as.numeric(l$l19_pnd_sold_alive)

l$l20_pnd_kept <- as.numeric(l$l20_pnd_kept)

l$l19_pnd_sold_alive <- as.numeric(l$l21_pnd_other)
```
# Population check
## Individual farms were balance is not 0 (death+non.death(slaughter+sold+kept)). Please check this 16 surveys:
i.e. Owor Gedge at the start of the outbreak had 20 pigs, no deaths but kept 2, maybe he kept 20 instead of 2. The balance is 18. 20-2=18
Apiyo Grace at the start of the outbreak had 1 pig, no deaths, he kept 10, maybe he kept 1 instead of 10. Balance is -9. 1-10=-9.
```{r eval=FALSE, include=FALSE}
datatable(l %>% 
  # group_by(l4_Sub_village) %>% 
  group_by(l2_Way_point, l6_Name_producer) %>% 
  summarise(Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            # Affected=sum(l15_N_affected, na.rm = TRUE),
            Not_dead=sum(l18_pnd_slaughtered, 
                         l19_pnd_sold_alive, 
                         l20_pnd_kept, na.rm = TRUE),
            Balance=Initial_pop-Deaths-Not_dead,
            # start=Initial_pop - Now,
    slaughther=sum(l18_pnd_slaughtered, na.rm = TRUE),
            Sold=sum(l19_pnd_sold_alive, na.rm = TRUE),
            Kept=sum(l20_pnd_kept, na.rm = TRUE),
            Interview=sum(l17_N_interview, na.rm = TRUE),
) %>% 
  arrange(desc(Balance)) %>% 
  filter(Balance != 0) %>% 
  print(n=100))


```

## Individual farms were balance is 0 (death+non.death(slaughter+sold+kept)). 83 entries are ok. Balance is checked considering the number of initial pigs at the beggining of the outbreak minus the slaughtered, sold or kept.
```{r}
datatable(l %>% 
  # group_by(l4_Sub_village) %>% 
  group_by(l2_Way_point) %>% 
  summarise(n(),Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            # Affected=sum(l15_N_affected, na.rm = TRUE),
            Not_dead=sum(l18_pnd_slaughtered, 
                         l19_pnd_sold_alive, 
                         l20_pnd_kept, na.rm = TRUE),
            Balance=Initial_pop-Deaths-Not_dead,
            # start=Initial_pop - Now,
    slaughther=sum(l18_pnd_slaughtered, na.rm = TRUE),
            Sold=sum(l19_pnd_sold_alive, na.rm = TRUE),
            Kept=sum(l20_pnd_kept, na.rm = TRUE),
            # Interview=sum(l17_N_interview, na.rm = TRUE),
) %>% 
  arrange(desc(Balance)) %>% 
  filter(Balance == 0) %>% 
  print(n=100))
```
## Balance check grouped by village.
```{r}
datatable(l %>% 
  group_by(l4_Sub_village) %>%
  # group_by(l6_Name_producer) %>% 
  summarise(n(),Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            # Affected=sum(l15_N_affected, na.rm = TRUE),
            Not_dead=sum(l18_pnd_slaughtered, 
                         l19_pnd_sold_alive, 
                         l20_pnd_kept, na.rm = TRUE),
            Balance=Initial_pop-Deaths-Not_dead,
            # start=Initial_pop - Now,
    slaughther=sum(l18_pnd_slaughtered, na.rm = TRUE),
            Sold=sum(l19_pnd_sold_alive, na.rm = TRUE),
            Kept=sum(l20_pnd_kept, na.rm = TRUE),
            # Interview=sum(l17_N_interview, na.rm = TRUE),
) %>% 
  arrange(desc(Balance)) %>% 
  # filter(Balance == 0) %>% 
  print(n=100))
```

## Calculating aparent prevalence 
```{r}
l$l61_apparent_prevalence <- l$l16_N_deaths/l$l14_N_init
l$l62attack_rate <- l$l16_N_deaths/l$l14_N_init

l %>% 
  select(l14_N_init, 
         l15_N_affected, 
         l16_N_deaths, 
         l18_pnd_slaughtered, 
         l19_pnd_sold_alive, 
         l20_pnd_kept,
         l17_N_interview,
         l61_apparent_prevalence) %>% 
  descr()

summary(l$l61_apparent_prevalence)
```
```{r}
l %>% 
  group_by(l4_Sub_village) %>% 
  select(l14_N_init, 
         l15_N_affected, 
         l16_N_deaths, 
         l18_pnd_slaughtered, 
         l19_pnd_sold_alive, 
         l20_pnd_kept,
         l17_N_interview,
         l61_apparent_prevalence) %>% 
  descr()
```

# Apparent prevalence by farm
```{r}
datatable(l %>% 
  # group_by(l4_Sub_village) %>%
  group_by(l2_Way_point) %>%
  summarise(n(), Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            # Affected=sum(l15_N_affected, na.rm = TRUE),
            Apparent_prevalence=round((Deaths/Initial_pop),4)*100) %>% 
  arrange(desc(Apparent_prevalence)) %>% 
  # filter(Balance == 0) %>% 
  print(n=100))


table(l$cases)

datatable(l %>% 
  # group_by(l4_Sub_village) %>%
  group_by(cases) %>%
  summarise(n(), 
            Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            Apparent_prevalence=round((Deaths/Initial_pop),4)*100) %>% 
    arrange(desc(Apparent_prevalence)))


library(epiR)
epi.prev(l$l16_N_deaths[l$cases == "Case"], l$l14_N_init[l$cases == "Case"], 0.95, 0.95, conf.level = 0.99)
```

# Total Apparent prevalence by cases and no cases
```{r}
datatable(l %>% 
  # group_by(l4_Sub_village) %>%
  group_by(cases) %>% 
  summarise(n(), Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            interview=sum(l17_N_interview, na.rm = TRUE),
            reduction=Initial_pop-interview,
            percent_red=(1-(interview/Initial_pop))*100,
            Apparent_prevalence=round((Deaths/Initial_pop),4)*100) %>% 
  arrange(desc(Apparent_prevalence)) %>% 
  # filter(Balance == 0) %>% 
  print(n=100))

```


# Total Apparent prevalence by village
```{r}
datatable(l %>% 
  # group_by(l4_Sub_village) %>%
  group_by(cases, l4_Sub_village) %>% 
  summarise(n(), Initial_pop=sum(l14_N_init, na.rm = TRUE),
            Deaths=sum(l16_N_deaths, na.rm = TRUE),
            interview=sum(l17_N_interview, na.rm = TRUE),
            reduction=Initial_pop-interview,
            percent_red=(1-(interview/Initial_pop))*100,
            Apparent_prevalence=round((Deaths/Initial_pop),4)*100) %>% 
  arrange(desc(Apparent_prevalence)) %>% 
  # filter(Balance == 0) %>% 
  print(n=100))

```

## Distribution of initial and final population over villages
```{r}
ggplot(l, aes(l14_N_init, cases, fill=l4_Sub_village))+
  geom_boxplot()+
  labs(
    y=NULL,
    x="Number of pigs at the begining of the outbreak"
  )

ggplot(l, aes(l17_N_interview, cases, fill=l4_Sub_village))+
  geom_boxplot()+
  labs(
    y=NULL,
    x="Number of pigs at the interview"
  )

```

# Comparison between apparent prevalence
```{r}
l$ap <- round((l$l16_N_deaths/l$l14_N_init),4)*100

table(l$ap)

ggplot(l, aes(ap, cases, fill=l4_Sub_village))+
  geom_boxplot()+
  labs(
    y=NULL,
    x="Apparent prevalence"
  )

library(forcats)

l <- l %>%
  group_by(l4_Sub_village) %>%
  mutate(median_ap = median(ap)) %>%
  ungroup() %>%
  mutate(l4_Sub_village = fct_reorder(l4_Sub_village, median_ap, .desc = TRUE))

# Plot the boxplot
ggplot(l, aes(x = l4_Sub_village, y = ap, fill = cases)) +
  geom_boxplot() +
  labs(
    y = NULL,
    x = "Apparent Prevalence"
  ) +
   theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
# Comparison between population modification
```{r}
l$p <- l$l14_N_init - l$l17_N_interview
summary(l$p)

l$pp <- round((l$l14_N_init+0.001)/l$l17_N_interview,5)*100
summary(l$pp)

ggplot(l, aes(ap, l4_Sub_village, fill=l4_Sub_village))+
  geom_boxplot()+
  labs(
    y=NULL,
    x="Apparent prevalence"
  )

library(forcats)

l <- l %>%
  group_by(l4_Sub_village) %>%
  mutate(median_ap = median(ap)) %>%
  ungroup() %>%
  mutate(l4_Sub_village 
         = fct_reorder(l4_Sub_village, median_ap, .desc = TRUE))

# Plot the boxplot
ggplot(data=l[l$cases == "Case",], 
       aes(x = l4_Sub_village, 
           y = ap, 
           fill = cases)) +
  geom_boxplot() +
  labs(
    y = NULL,
    x = "Apparent Prevalence") +
   theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# How many pigs did the household have at the start of the outbreak


# What did you do with the pigs that died?
## Check errors of problems (nothing wrong found)
```{r}
l$l23_pd_Ate[l$l23_pd_Ate == "NA"] <- NA
l$l24_pd_Sold[l$l24_pd_Sold == "NA"] <- NA
l$l25_pd_buried[l$l25_pd_buried == "NA"] <- NA
l$l26_pd_burned[l$l26_pd_burned == "NA"] <- NA
l$l27_pd_latrine[l$l27_pd_latrine == "NA"] <- NA
l$l28_pd_bush_swamp_river[l$l28 == "NA"] <- NA

l %>% 
  filter(is.na(l23_pd_Ate) & l16_N_deaths != 0)

l %>% 
  filter(is.na(l24_pd_Sold) & l16_N_deaths != 0)

l %>% 
  filter(is.na(l25_pd_buried) & l16_N_deaths != 0)

l %>% 
  filter(is.na(l26_pd_burned) & l16_N_deaths != 0)

l %>% 
  filter(is.na(l27_pd_latrine) & l16_N_deaths != 0)

```
## Changing to numerical to quantify
```{r}
l$pd_ate <- l$l23_pd_Ate
l$pd_ate <- gsub("Yes", "1", l$pd_ate)
l$pd_ate <- gsub("No", "0", l$pd_ate)

l$pd_sold <- l$l24_pd_Sold
l$pd_sold <- gsub("Yes", "1", l$pd_sold)
l$pd_sold <- gsub("No", "0", l$pd_sold)

l$pd_buried <- l$l25_pd_buried
l$pd_buried <- gsub("Yes", "1", l$pd_buried)
l$pd_buried <- gsub("No", "0", l$pd_buried)

l$pd_burned <- l$l26_pd_burned
l$pd_burned <- gsub("Yes", "1", l$pd_burned)
l$pd_burned <- gsub("No", "0", l$pd_burned)

l$pd_latrine <- l$l27_pd_latrine
l$pd_latrine <- gsub("Yes", "1", l$pd_latrine)
l$pd_latrine <- gsub("No", "0", l$pd_latrine)

l$pd_bush_river <- l$l28_pd_bush_swamp_river
l$pd_bush_river <- gsub("Yes", "1", l$pd_bush_river)
l$pd_bush_river <- gsub("No", "0", l$pd_bush_river)

l$pd_other <- l$l29_pd_other
l$pd_other <- gsub("Yes", "1", l$pd_other)
l$pd_other <- gsub("No", "0", l$pd_other)

l$pd_ate <- as.numeric(l$pd_ate)
l$pd_sold <- as.numeric(l$pd_sold)
l$pd_buried <- as.numeric(l$pd_buried)
l$pd_burned <- as.numeric(l$pd_burned)
l$pd_latrine <- as.numeric(l$pd_latrine)
l$pd_bush_river <- as.numeric(l$pd_bush_river)
l$pd_other <- as.numeric(l$pd_ate)
```


```{r}
actions_sick <- data.frame(t(l %>% 
  group_by(cases) %>% 
  summarise(
            Ate=sum(pd_ate, na.rm =TRUE),
            Sold=sum(pd_sold, na.rm =TRUE),
            Burried=sum(pd_buried, na.rm =TRUE),
            Burned=sum(pd_burned, na.rm =TRUE),
            Latrine=sum(pd_latrine, na.rm =TRUE),
            Bush=sum(pd_bush_river, na.rm =TRUE),
            Other=sum(pd_other, na.rm =TRUE),)))

actions_sick$resposes <- as.numeric(actions_sick$X2)/40
datatable(actions_sick)

l %>% 
  group_by(l$l4_Sub_village) %>%
  summarise(Ate=sum(pd_ate, na.rm =TRUE),
            Sold=sum(pd_sold, na.rm =TRUE),
            Burried=sum(pd_buried, na.rm =TRUE),
            Burned=sum(pd_burned, na.rm =TRUE),
            Latrine=sum(pd_latrine, na.rm =TRUE),
            Bush=sum(pd_bush_river, na.rm =TRUE),
            Other=sum(pd_other, na.rm =TRUE),)

l <- l %>% 
  mutate(individual_what_did_sick_pigs=sum(pd_ate,
                                           pd_sold, 
                                           pd_buried,
                                           pd_burned,
                                           pd_latrine,
                                           pd_bush_river,
                                           pd_other, na.rm = TRUE))

table(l$individual_what_did_sick_pigs)

table(l$cases)
```

# Map
## Checking coordinates
Preparing coordinates

## ? 31	Check Acayo Nighthy Santo nort coordinate (y), it is the same as x. Correct.
## ? 62 Check Christine Lukioiya / Innocen atimango easting (x) coordinate

```{r}
l$y <- l$l9_Norting
l$y <- gsub("-", ".", l$y)
l$y <- as.numeric(l$y)

l$x <- l$l10_Easting
l$x <- gsub("-", ".", l$x)
l$x <- as.numeric(l$x)

#? coordinates
l$y[l$l2_Way_point == "31"] <- "2.57189"
l$x[l$l2_Way_point == "56"] <- "32.061810"

# plot(l$x, l$y)

```

## Manipulable map of the complete dataset
```{r}
library(leaflet)
library(stringr)
s <- "Alfredo Acosta"
paste(substr(word(s, 1), 1,3), substr(word(s, 2), 1,3))


l$owner <- paste(substr(word(l$l6_Name_producer, 1), 1,3), substr(word(l$l6_Name_producer, 2), 1,2))
l$id <- l$l2_Way_point
l$village <- l$l4_Sub_village
l$x <- as.numeric(l$x)
l$y <- as.numeric(l$y)

map <- leaflet(l) %>%
  addProviderTiles(providers$Esri.WorldImagery)

# palete for case and control
pal <- colorFactor(c("red", "blue"), domain=c("Case", "Control"))

mc <- map %>%
  setView(lng = 32, lat = 2, zoom = 7) %>%
  addCircleMarkers(~x, ~y,
            stroke = FALSE, fillOpacity = 0.4,
             color = ~pal(cases),
             popup = ~paste("ID:", id,
                            "<br>", "Owner:", owner,
                             "<br>", "Prevalence", ap,
                             "<br>", "Deaths", l16_N_deaths))
```


## Manipulable map of the cases (farms with reported deaths)
```{r}

library(viridis)
library(leafsync)

map <- leaflet(l) %>%
  addProviderTiles(providers$OpenStreetMap.Mapnik)

# palete for case and control
pal <- colorFactor(viridis_pal(option = "D")(length(unique(l$l4_Sub_village))), 
                   domain=unique(l$l4_Sub_village))

mv <- map %>%
  setView(lng = 32, lat = 2, zoom = 7) %>%
  addCircleMarkers(~x, ~y,
            stroke = FALSE, fillOpacity = 1,
             color = ~pal(l4_Sub_village),
             popup = ~paste("Village", l4_Sub_village, 
                            "<br>","ID:", id,
                            # "<br>", "Owner:", owner,
                             "<br>", "Gender", l7_Gender,
                             "<br>", "pigs_interv", l17_N_interview)) %>% 
    addLegend(pal = pal,
            values = ~l4_Sub_village,
    position = "bottomleft")
```
# Combinaed map cases and villages
```{r}
leafsync::sync(mc,mv)

```


## pub map Lutuk
```{r}
library(ggmap)
# lu <- get_stadiamap(bbox = c(left = 31, 
#                                bottom = 2, 
#                                right = 32, 
#                                top = 3),
#                       zoom = 8, 
#                       arg = "terrain-background")

```

# Spatial temporal analysis
## Preparing satscan files
```{r}

# pop
# DPA_parr, amount, fecha
# 
# cases
# DPA_parroq, amount, date
# 
# geo
# DPA_parroq, lon, lat


pop <- l %>% 
  select(l2_Way_point, l14_N_init, Date_first_detection_ill_pig) %>%
  mutate(Date_first_detection_ill_pig_completed=ifelse(is.na(Date_first_detection_ill_pig), as.Date("2024-01-01"), Date_first_detection_ill_pig)) %>% 
  group_by(l2_Way_point, l14_N_init, Date_first_detection_ill_pig, Date_first_detection_ill_pig_completed)

pop$Date_first_detection_ill_pig_completed <- as.Date(pop$Date_first_detection_ill_pig_completed)

# First run with cases of october
cases <- l %>% 
  filter(cases == "Case") %>% 
  filter(!is.na(Date_first_detection_ill_pig)) %>% 
  select(l2_Way_point, l16_N_deaths, Date_first_detection_ill_pig) %>% 
  group_by(l2_Way_point, l16_N_deaths, Date_first_detection_ill_pig)

#second run with the cases since 2024-01-01
cases <- l %>% 
  filter(cases == "Case") %>% 
  filter(year(Date_first_detection_ill_pig) == "2024") %>% 
  filter(!is.na(Date_first_detection_ill_pig)) %>% 
  select(l2_Way_point, l16_N_deaths, Date_first_detection_ill_pig) %>% 
  group_by(l2_Way_point, l16_N_deaths, Date_first_detection_ill_pig)


geo <- l %>% 
  select(l2_Way_point, y, x) %>% 
  group_by(l2_Way_point, y, x)

# write.csv(pop, file = "pop.csv")
# write.csv(cases, file = "cases.csv")
# write.csv(geo, file = "geo.csv")

```

# Credits 
Acosta, Alfredo PhD<sup>1</sup>. 
**SVA<sup>1</sup>**: SVA <http://www.sva.se/>. 